---
title: "Installing Bio3D"
author: "Xin-Qiu Yao, Lars Skjaerven & Barry J. Grant"
date: "September 25, 2016"
affiliation: University of Bergen (Norway), and University of Michigan (Ann Arbor, US)
output:
  pdf_document:
    highlight: default
    keep_tex: yes
    number_sections: yes
    toc: no
    fig_width: 5
    fig_height: 5
    fig_caption: yes
  md_document:
    fig_height: 5
    fig_width: 5
    variant: markdown_phpextra
documentclass: article
fontsize: 11pt
geometry: tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm
linkcolor: black
---

<br><br>

# Quick Install  

On all platforms (Mac, Linux, and PC) open R (version 3.1.0 or higher) and use the function 
**install.packages()** at the R command prompt to install the package from your 
local [CRAN](http://cran.r-project.org/web/packages/bio3d/) site:

```{r, eval=FALSE}
install.packages("bio3d", dependencies=TRUE)
```

Alternatively, if you prefer to use [RStudio](https://www.rstudio.com/products/rstudio/), select 
`Tools` $\rightarrow$ `Install Packages..`, Enter **bio3d** in the `Packages` text box (and make 
sure `Install Dependencies` is selected). Then click `Install`.



## Optional Extras
For full Bio3D functionality you should have *MUSCLE*, *DSSP*, and 
*NetCDF* (headers and libraries) installed on your system and in the search path 
for executables. Background details and install instructions for each of these packages on different 
operating systems is provided further below. 

For quick install on a **Mac** we recommend using [homebrew](http://brew.sh/index.html):

    brew install homebrew/science/muscle
    brew install homebrew/science/netcdf
    brew install homebrew/science/dssp

On a **Linux/UNIX** system you should use your appropriate package manager (e.g. *apt-get* for Debian/Ubuntu and *dnf* for Red Hat/Fedora systems, see below). If you experience problems with any of these steps please read on for alternative installation methods.


## Testing your installation

You should now be able to load the Bio3D package into your current R
session by typing the usual `library(bio3d)` command at the R Console.

```{r, eval=FALSE}
library(bio3d) 
help(package="bio3d")
vignette(package="bio3d") 
```

We now suggest you use the command `demo("pdb")`, `demo("pca")` and
`demo("md")` to get a quick feel for some of the tasks that we will be
introducing in subsequent vignettes:

```{r, eval=FALSE}
library(bio3d) 
demo("pdb") 
demo("pca") 
demo("md") 
```
<br>


# Detailed Installation Instructions

Before you attempt to install Bio3D you should have a relatively recent
version of R installed and working on your system (we recommend at least
**R version 3.1.0**). Detailed instructions for obtaining and installing R
on various platforms can be found on the R home page <http://www.r-project.org>.

Bio3D makes use of a number of **additional R packages** including *ncdf*,
*bigmemory*, *XML*, *RCurl* and *igraph*.
Make sure required headers and libraries are installed on your system prior to installing
these R packages. i.e. the R package `ncdf` requires `netcdf` and 
`libnetcdf-dev`; `XML` requires `libxml2-dev`; and `RCurl` requires `libcurl3-dev`. 

In addition, for full Bio3D functionality you should have *MUSCLE* and *DSSP*
installed on your system and in the search path for executables. 


**Note**: If you encounter difficulties in installing any of the suggested packages, note
that most functions in Bio3D will work just fine without. e.g. `ncdf` is only required
for reading/writing netcdf binary trajectory files; `igraph` for visualization of 
correlation network analysis; `RCurl` and `XML` for searching and fetching data from 
various biomolecular databases. 


## Obtaining Bio3D

The Bio3D package is available in two forms from
[CRAN](http://cran.r-project.org/web/packages/bio3d/) and 
the [Bio3D website](http://thegrantlab.org/bio3d/)

-   [as platform independent source
    code](http://thegrantlab.org/bio3d/download) (intended
    primarily for Mac and Unix systems),

-   [as a compiled binary for
    Windows](http://thegrantlab.org/bio3d/download).

To install from source requires that your machine has standard compilers
and tools such as Perl 5.004 or later. If you run into problems with
source installation please refer to section 6.1 of the [R Installation
and Administration Manual](http://cran.r-project.org/doc/manuals/R-admin.html). 
Below we provide installation instructions for some of the most common OS's. 


## MacOS X Installation

R on Mac OS X can be used either on the command-line, like on other UNIX
systems, via the R.app GUI (included with your binary R install), or the increasingly popular [RStudio](https://www.rstudio.com/products/rstudio/) IDE. 

Regardless of your preferred interface you should be able to find the R command prompt and 
install the Bio3D package from [CRAN](http://cran.r-project.org/web/packages/bio3d/) using 
the following command:

```{r, eval=FALSE}
install.packages("bio3d", dependencies=TRUE)
```

We recommend using the [homebrew](http://brew.sh/index.html) package manager for installation 
of MUSCLE, DSSP and NetCDF. For more information on homebrew see <http://brew.sh/index.html>. 
Using homebrew these packages can be installed with the following command:

      brew install homebrew/science/muscle
      brew install homebrew/science/netcdf
      brew install homebrew/science/dssp
    

Alternatively, you can use the `Packages and Data` menu of the R.app GUI, in
particular the sub-item Package Installer: Download the source tar.gz
file from above. In the R GUI select `Packages and Data` $\rightarrow$
`Package Installer` $\rightarrow$ `Local Source Package`, and press the
`Install` button. Select the Bio3D tar file and press `Open`.

In RStudio, select `Tools` $\rightarrow$ `Install Packages..`, Enter **bio3d** in the 
`Packages` text box, make sure `Dependencies` is selected and click `Install`.



## Linux Systems (Ubuntu & Fedora)
On a Debian system (e.g. Ubuntu) most required packages and programs
can be installed directly through the official package manager system 
with the `apt-get install` command:

    apt-get install r-base-core netcdf-bin libnetcdf-dev libxml2-dev \
                    libcurl3-dev seaview muscle pymol 

For Red Hat based systems (e.g. Fedora) the equivalent command is:

    dnf install R-base R-devel netcdf-devel netcdf libxml2-devel \
                libcurl-devel seaview pymol 

The Bio3D package can be obtained and installed via [CRAN](http://cran.r-project.org/web/packages/bio3d/). 
Start R by issuing the command `R` and then from the R prompt install the Bio3D package:

```{r, eval=FALSE}
install.packages("bio3d", dependencies=TRUE)
```

Alternatively, Bio3D can be downloaded as source code e.g.
from <https://bitbucket.org/Grantlab/bio3d/downloads> and installed
with the command:

```{r, eval=FALSE}
install.packages("bio3d_2.3-0.tar.gz")
```

Note that **MUSCLE** is not available from the Fedora package manager, but can be installed by:

    wget http://www.drive5.com/muscle/downloads3.8.31/muscle3.8.31_i86linux64.tar.gz
    tar xzvf muscle3.8.31_i86linux64.tar.gz
    mv muscle3.8.31_i86linux64 /usr/local/bin/muscle
    chmod a+x /usr/local/bin/muscle

**DSSP** is also not available from a number of the package managers, but can be installed by:

    wget ftp://ftp.cmbi.ru.nl/pub/software/dssp/dssp-2.0.4-linux-amd64 -O /usr/local/bin/dssp
    chmod a+x /usr/local/bin/dssp


### Installing a minimal version of Bio3D {.unnumbered}
A minimal version of Bio3D with reduced functionality (i.e. for reading/writing
binary trajectory files, and fetching data from various databases)
will require only the R base installed (i.e. no additional packages needed).
Thus, only the R base will be required. In Ubuntu this can be obtained with the
following two commands:

    apt-get install r-base-core

```{r, eval=FALSE}
install.packages("bio3d", dependencies=FALSE)
```

### Installing required R packages individually {.unnumbered}
The Bio3D dependencies can be installed from within R with the command `install.packages`:

```{r, eval=FALSE}
# install only the XML package
install.packages("XML")

# install all required
install.packages(c("XML", "RCurl", "ncdf", "igraph", "bigmemory"), 
                 dependencies=TRUE)
```



## Windows Installation

To install the Bio3D package on Windows download the compiled binary
.zip file from above.

Start R and from GUI click `Packages` $\rightarrow$
`Install Package(s) from local zip file` then simply select your
downloaded Bio3D zip file and click `Open` to finish the installation.



## Installing the development version of Bio3D

For the majority of users we recommend the use of the last stable
release available from the [main Bio3D
website](http://thegrantlab.org/bio3d/download). The
development version is available from our [bitbucket
repository](https://bitbucket.org/Grantlab/bio3d/) and typically
contains new functions and bug fixes that have not yet been incorporated
into the latest stable release.

There are several ways to download and install the development version
of Bio3D. The simplest method is to install directly from our bitbucket
repository using the R function `install_bitbucket()` from the
`devtools` package.

```{r, eval=FALSE}
install.packages("devtools")
library(devtools)
install_bitbucket("Grantlab/bio3d", subdir = "ver_devel/bio3d/")
```

Alternative installation methods and additional instructions are posted
to the wiki section of our [bitbucket
repository](https://bitbucket.org/Grantlab/bio3d/).
<br><br>



# Additional utilities

There are a number of additional packages and programs that will either
interface directly with Bio3D (MUSCLE, DSSP and STRIDE), or that we
consider generally invaluable for working with biomolecular structure
and sequence data (e.g. VMD, PyMOL, and SEAVIEW). A brief description of
how to obtain these additional packages is given below.

## Required for full Bio3D functionality

### MUSCLE: {.unnumbered} 

Muscle is a fast multiple sequence alignment program available from the
muscle home page <http://www.drive5.com/muscle>. The Bio3D functions
`seqaln()` and `pdbaln()` currently calls the MUSCLE program, hence
MUSCLE must be installed on your system and in the search path for
executables if you wish to use this function.

**A note for Mac and Unix users**: 

After downloading MUSCLE, it should be unzipped and renamed to just
“muscle” and placed in a directory such as “/usr/local/bin/” (i.e. in your PATH).

### DSSP: {.unnumbered} 

DSSP a popular secondary structure analysis program which should be installed
on your system as an executable called “dssp” or "mkdssp" and be in the search path
for executables. DSSP is available from a number of sources including:

-   <http://www.cmbi.ru.nl/dssp.html>

-   <http://swift.cmbi.ru.nl/gv/dssp/DSSP_5.html>.

## Optional

### STRIDE: {.unnumbered} 

STRIDE is another secondary structure analysis program available from the
[EMBL-Heidelberg](http://webclu.bio.wzw.tum.de/stride/). Stride is
similar in functionality to the more prevalent DSSP (see above).
However, stride is often much easier to setup on different computer
systems as you may be able to simply copy or link to the stride
executable distributed within every version of VMD (see below).

### SEAVIEW: {.unnumbered} 

SEAVIEW is a graphical multiple sequence alignment editor. Download
information and documentation are available from PBIL
<http://pbil.univ-lyon1.fr/software/seaview.html>. I use Seaview to
manually check and edit protein sequence alignment files prior to
detailed analysis. I believe this should be done with every alignment
regardless of how accurate the various automatic tools are supposed to
be.

### Clustal Omega: {.unnumbered} 

Clustal Omega is multiple sequence alignment program that can be used as 
an alternative to MUSCLE (needed e.g. for functions `seqaln()` and `pdbaln()`). 
Clustal Omega is available from <http://www.clustal.org/omega/>.

### VMD: {.unnumbered} 

VMD is a molecular visualization program for displaying, animating, and
analyzing large biomolecular systems using 3-D graphics. Visit the VMD
website for download information and documentation
<http://www.ks.uiuc.edu/Research/vmd/>.

### PyMOL: {.unnumbered} 

PyMOL is another visualization program. Bio3D functions `pymol.dccm()` and `pymol.modes()` require PyMOL
to be in your search path. PyMOL is available from
<http://www.pymol.org>.
<br><br>


# A Note on Calling External Programs from R/Bio3D
Ideally, as mentioned previouly, MUSCLE and DSSP should installed on your system and be in the search path for executables. To test this you should be able to call these programs from the command line with just their name from any directory.

For **Mac and Linux** you can find out whats in your **PATH** by launching your favorate Terminal program (on Mac one called Terminal can be found in Applications/Utilities folder) and entering:

    echo $PATH

And the result should be like this…

    /usr/bin:/bin:/usr/sbin:/sbin:/usr/local/bin

So this is stating that you can run Unix style applications located in 5 default locations of path in the file system:

    /usr/bin
    /bin
    /usr/sbin
    /sbin
    /usr/local/bin

You can add extra locations to your path by creating or editing an existing **.bash_profile** file in your home directory. This file should contain a line like the following:

    export PATH="/my/new/path:$PATH"  

You can now put muscle and dssp in any of the locations listed by echo $PATH, including /my/new/path/, which of course you should change to something sensible for you.

For **Windows**, right click "My computer" -> click "Change settings" -> Advanced -> Environment Variables -> From "System variables" list find "Path" and click "Edit" -> Add the path to your programs at the end of the line. 


# Where to next

If you have read this far, congratulations! We are ready to have some
fun and move on to other package vignettes that describe various analysis
including Molecular Dynamics Trajectory Analysis, Correlation
Network Analysis (where we will build and dissect dynamic networks form
different correlated motion data), enhanced methods for Normal Mode
Analysis (where we will explore the dynamics of large protein families
and superfamilies), and advanced Comparative Structure Analysis (where
we will mine available experimental data and supplement it with
simulation results to map the conformational dynamics and coupled
motions of proteins). Happy Bio3Ding!
<br><br>

# Session Information

The version number of R and packages loaded for generating the vignette
were: 
```{r}
sessionInfo() 
```


[^1]: The latest version of the package, full documentation and further
    vignettes (including detailed installation instructions) can be
    obtained from the main Bio3D website:
    [http://thegrantlab.org/bio3d/](http://thegrantlab.org/bio3d/)

[^2]: This vignette contains executable examples, see `help(vignette)` for further details.



