---
title: "Getting started with Bio3D"
author: "Lars Skjaerven, Xin-Qiu Yao & Barry J. Grant"
date: "September 28, 2014"
affiliation: University of Michigan, Ann Arbor
output:
  pdf_document:
    highlight: default 
    keep_tex: yes
    number_sections: yes
    toc: yes
    fig_width: 5
    fig_height: 5
    fig_caption: yes
  md_document:
    fig_height: 5
    fig_width: 5
    variant: markdown_phpextra
bibliography: Bio3D_pdb.bib
documentclass: article
fontsize: 11pt
geometry: tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm
linkcolor: black
---

# Introduction

Bio3D[^1] is an R package containing utilities for the analysis of biomolecular
structure, sequence and trajectory data [@grant06].  Features include the ability
to read and write biomolecular structure, sequence and dynamic trajectory data,
perform atom selection, re-orientation, superposition, rigid core
identification, clustering, distance matrix analysis, conservation
analysis, normal mode analysis and principal component analysis. Bio3d
takes advantage of the extensive graphical and statistical capabilities of the 
[R environment](http://www.r-project.org) 
and thus represents a useful framework for exploratory analysis of structural data.


## Using this vignette
The aim of this document, termed a vignette[^2] in R parlance, is to provide a 
brief overview of Bio3D. A number of other Bio3D package vignettes are
available, including:

At the time of writing these include:

-   `Installing Bio3D`

-   `Comparative protein structure analysis with Bio3D`

-   `Sequence conservation analysis with Bio3D`

-   `Beginning trajectory analysis with Bio3D`

-   `Dynamic network analysis with Bio3D`

-   `Normal mode analysis with Bio3D`



# Getting started
Start R (type R at the comand prompt or, on Windows, double click on the R icon) 
and load the bio3d package by typing `library(bio3d)` at the R console prompt. 

```{r eval=TRUE, echo=FALSE}
library(bio3d)
```

Then use the command `lbio3d()` and `help()` to list the functions within the package
```{r eval=FALSE}
lbio3d()
help(bio3d)
```

## Finding Help
To get help on a particular function try `?`function or `help(function)`. 
For example, `?pca.xyz`

```{r eval=FALSE}
?pca.xyz
``` 

To search the help system for documentation matching a particular word or topic use the 
command `help.search("topic")`. For example, `help.search("pdb")`

```{r eval=FALSE}
help.search("pdb")
 ```

Typing `help.start()` will start a local HTML interface. 
After initiating `help.start()` in a session the `?function` commands will open as HTML pages. 
To execute examples for a particular function use the command `example(function)`. 
To run examples for the `read.dcd` function try example(`read.dcd`) 


## Bio3d Demo
Run the command `demo(bio3d)` to obtain a quick overview.

```{r eval=FALSE}
demo(bio3d)
``` 

The bio3d package consists of input/output functions, conversion and manipulation functions, analysis functions, and graphi
cs functions all of which are fully documented. Remember that you can get help on any particular function by using the comm
and ?function or help(function) from within R. 

```{r eval=FALSE}
help(read.pdb)
```


## Example Function Usage}
To better understand how a particular function operates it is often helpful to view and execute an example. Every function within the bio3d package is documented with example code that you can view by issuing the help command. 

Running the command example(function) will directly execute the example for a given function. 
In addition, a numer of worked examples are available as short Tutorials on the bio3d wiki. 

```{r eval=FALSE}
example(plot.bio3d)
```



# Working with PDB files


## Read a PDB file


```{r eval=TRUE}
pdb <- read.pdb("1atp")

print(pdb)

print(pdb$xyz)
```


```{r eval=TRUE}
# Print data for the first five atoms
pdb$atom[1:5, ]

# print residue numbers for the first atoms
head(pdb$atom$resno)
     
# Print some coordinate data
head(pdb$atom[1:5, c("x","y","z")])
     
# Or coordinates as a numeric vector
head(pdb$xyz)
```

## Atom selection 

```{r eval=TRUE}
# select all C-alpha atoms
ca.inds <- atom.select(pdb, "calpha")

print(ca.inds)

# selected atoms
head(pdb$atom[ca.inds$atom, ])

# select chain I
i.inds <- atom.select(pdb, chain="I")

print(i.inds)

# select C-alphas of chain I
cai.inds <- atom.select(pdb, "calpha", chain="I")

# coordinates of selection 
pdb$xyz[,cai.inds$xyz]

```

## Manipulate a PDB object

```{r eval=TRUE}
# generate C-alpha only PDB
ca.pdb <- trim.pdb(pdb, ca.inds)

# write C-alpha PDB to disk
write.pdb(ca.pdb, file="1ATP_CA.pdb")

# use chain A and B instead of E and I
e.inds <- which(ca.pdb$atom$chain == "E")
i.inds <- which(ca.pdb$atom$chain == "I")

ca.pdb$atom$chain[ e.inds ] <- "A"
ca.pdb$atom$chain[ i.inds ] <- "B"

# renumber residues
ca.pdb$atom$resno[e.inds] <- seq(1, length(e.inds))
ca.pdb$atom$resno[i.inds] <- seq(1, length(i.inds))
```

## Temperature factor analysis

```{r eval=TRUE}
plot.bio3d(ca.pdb$atom$b, sse=ca.pdb)

```


## Simiple structural alignment
```{r eval=TRUE}
# read two PKA structures
a <- read.pdb("1atp")
b <- read.pdb("1cmk")

# different sequence 
#a
#b

# perform iterative alignment
aln <- struct.aln(a, b)

# store new coordinates of protein B
b$xyz <- aln$xyz

```

## Concatenate multiple PDBs
```{r eval=TRUE}

a1 <- trim.pdb(a, atom.select(a, chain="E"))
b1 <- trim.pdb(b, atom.select(b, chain="E"))

a2 <- trim.pdb(a, atom.select(a, chain="I"))
b2 <- trim.pdb(b, atom.select(b, chain="I"))

# concatenate PDBs
new <- cat.pdb(a1, b1, a2, b2)
unique(new$atom$chain)

# write new PDB object to file
write.pdb(new, file="1ATP-1CMK.pdb")

```


# Document Details {.unnumbered}
This document is shipped with the Bio3D package in both R and PDF formats. All code
can be extracted and automatically executed to generate Figures and/or the PDF with the following commands:

```{r eval=FALSE}
library(rmarkdown)
render("Bio3D_pdb.Rmd", "all")
```


# Information About the Current Bio3D Session {.unnumbered}
```{r}
sessionInfo() 
```


# References {.unnumbered}



[^1]: The latest version of the package, full documentation and further
    vignettes (including detailed installation instructions) can be
    obtained from the main Bio3D website:
    [thegrantlab.org/bio3d/](http://thegrantlab.org/bio3d/).

[^2]: This vignette contains executable examples, see `help(vignette)`
    for further details.

[^3]: See also dedicated vignettes for *ensemble NMA* provided with the Bio3D package.