---
title: "Integrating normal modes and correlation network analysis in Bio3D"
author: "Xin-Qiu Yao, Guido Scarabelli, Lars Skjaerven & Barry J. Grant"
date: "October 1, 2014"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 5
    fig_width: 5
    highlight: default
    keep_tex: yes
    number_sections: yes
    toc: no
  md_document:
    fig_height: 5
    fig_width: 5
    variant: markdown_phpextra
documentclass: article
fontsize: 11pt
geometry: tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm
linkcolor: black
bibliography: Bio3D_cna-transducin.bib
affiliation: University of Bergen (Norway), and University of Michigan (Ann Arbor, US)
---

# Background  {.unnumbered}
Bio3D[^1] is an R package that provides interactive tools for structural bioinformatics. The primary focus of Bio3D is the analysis of bimolecular structure, sequence and simulation data [@grant06].

Correlation network analysis can be employed to identify protein segments with correlated motions.
In this approach, a weighted graph is constructed where each residue represents a node and the 
weight of the connection between nodes, *i* and *j*, represents their respective cross-correlation 
value, cij, expressed for example by the Pearson-like form or the linear mutual information. In this 
example, Normal mode analysis (NMA) is employed for the calculation of cross-correlations[^2].

## Requirements {.unnumbered}
Detailed instructions for obtaining and installing the Bio3D package on various platforms can be found in the [**Installing Bio3D Vignette**](http://thegrantlab.org/bio3d/tutorials) available both on-line and from within the Bio3D package. Note that this vignette also makes use of the [IGRAPH](http://igraph.org/) R package, which can be installed with the command

```{r eval=FALSE}
install.packages("igraph")
```

## About this document {.unnumbered}
This vignette was generated using **Bio3D version `r packageVersion("bio3d")`**.


# Correlation network analysis based on single-structure NMA
In this first example we perform *NMA* on two crystallographic structures of transducin G protein alpha subunits, an active
GTP-analog-bound structure (PDB id *1tnd*) and an inactive GDP- and 
GDI (GDP dissociation inhibitor)-bound structure (PDB id *1kjy*), by calling the function **nma()**.
Cross-correlation matrices are calculated with the function **dccm()**. Correlation networks are
then constructed with the function **cna()**. Network visualization is finally performed
with the function **plot.cna()**. See also `help(cna)` for more details and example analysis.



```{r}
library(bio3d)
```

First we read our selected GTP and GDI bound structures, select chain A, and perform NMA on each individually.
```{r single-pdb, results="hide"}
pdb.gtp = read.pdb("1TND")
pdb.gdi = read.pdb("1KJY")

pdb.gtp = trim.pdb(pdb.gtp, inds=atom.select(pdb.gtp, chain="A"))
pdb.gdi = trim.pdb(pdb.gdi, inds=atom.select(pdb.gdi, chain="A"))
```

```{r nma}
modes.gtp = nma(pdb.gtp)
modes.gdi = nma(pdb.gdi)
```


Residue cross-correlations can then be calculated based on these normal mode results.
```{r nma_cij, cache=TRUE, results="hide", warning=FALSE, message=FALSE}
cij.gtp = dccm(modes.gtp)
cij.gdi = dccm(modes.gdi)
```

Correlation networks for both conformational states are constructed by applying the
function **cna()** to the corresponding correlation matrices. Here, the C-alpha atoms
represents nodes which are interconnected by edges with weights corresponding to the pairwise correlation coefficient. 
Edges are only constructed for pairs of nodes which obtain a coupling strength larger than a specified cutoff value
(0.35 in this example). The weight of each edge is calculated as -log(|cij|), where cij is the 
correlation value between two nodes *i* and *j*. For each correlation network, betweenness clustering 
is performed to generate aggregate nodal clusters, or communities, that are highly 
intra-connected, but loosely inter-connected. By default, **cna()** returns communities 
associated with the maximal modularity value. 

```{r cna, results="hide", message=FALSE}
net.gtp = cna(cij.gtp, cutoff.cij=0.35)
net.gdi = cna(cij.gdi, cutoff.cij=0.35)
```

```{r display-cna}
net.gtp
net.gdi
```

A 3-D visualization of networks can also be performed with the Bio3D function **view.dccm()** (See `help(view.dccm)` and the vignette "Enhanced Methods for Normal Mode Analysis with Bio3D" available on-line):
```{r 3dview, eval=FALSE}
view.dccm(net.gtp$cij, launch=TRUE)
```


Maximization of modularity sometimes creates unexpected community partitions splitting 
natural structure motifs, such as secondary structure elements, into many small community 'islands'. To avoid this situation, we look into partitions with modularity close to the maximal value 
but with an overall smaller number of communities. 

```{r new-fun}
mod.select <- function(x, thres=0.005) {
   remodel <- community.tree(x, rescale = TRUE)
   n.max = length(unique(x$communities$membership))
   ind.max = which(remodel$num.of.comms == n.max)
   v = remodel$modularity[length(remodel$modularity):ind.max]
   v = rev(diff(v))
   fa = which(v>=thres)[1] - 1
   ncomm = ifelse(is.na(fa), min(remodel$num.of.comms), n.max - fa)

   ind <- which(remodel$num.of.comms == ncomm)
   network.amendment(x, remodel$tree[ind, ])
}

```{r modularity_selection, cache=TRUE}
nnet.gtp = mod.select(net.gtp)
nnet.gtp
nnet.gdi = mod.select(net.gdi)
nnet.gdi
```

The resulting networks can be visualized with the Bio3D function **plot.cna()**, which can generate 2D representations for both full residue-level and coarse-grained community-level networks.

```{r layout}
cent.gtp.full = layout.cna(nnet.gtp, pdb=pdb.gtp, full=TRUE, k=3)[,1:2]
cent.gtp = layout.cna(nnet.gtp, pdb=pdb.gtp, k=3)[,1:2]
cent.gdi.full = layout.cna(nnet.gdi, pdb=pdb.gtp, full=TRUE, k=3)[,1:2]
cent.gdi = layout.cna(nnet.gdi, pdb=pdb.gtp, k=3)[,1:2]
```

The following code plots the four networks. 

```{r figure1, fig.cap="Correlation networks for GTP \"active\" and GDI \"inhibitory\" conformational states of transducin. Networks are derived from NMA applied to single GTP and GDI crystallographic structures.", echo=c(3,5,6,8)}
layout(matrix(c(1:4), 2, 2))
par(mar=c(0.1, 0.1, 3, 0.1))
plot.cna(nnet.gtp, layout=cent.gtp.full, full=TRUE, vertex.label=NA, vertex.size=5)
title(main="GTP")
plot.cna(nnet.gtp, layout=cent.gtp)
plot.cna(nnet.gdi, layout=cent.gdi.full, full=TRUE, vertex.label=NA, vertex.size=5)
title(main="GDI")
plot.cna(nnet.gdi, layout=cent.gdi)
```


# Correlation network analysis based on ensemble NMA
In this example we perform ensemble NMA on 53 crystallographic transducin G-alpha structures, which can be 
categorized into active GTP-analog-bound, inactive GDP-bound, and inhibitory GDI-bound states. For this analysis
we utilize the example transducin structure dataset (further details of which can be obtained via `help(transducin)` along with the "Comparative Protein Structure Analysis with Bio3D" vignette available on-line). Briefly, this dataset includes aligned PDB coordinates (`pdbs`), structural invariant core positions (`core`) and annotations for each PDB structure (`annotation`).
Cross-correlation matrices for all structures in the ensemble are calculated with the function **dccm()**. 
State-specific ensemble average correlation matrices are then obtained with the function **dccm.filter()**.
Correlation networks are finally constructed with the function **cna()**, and visualization of the networks
performed with the function **plot.cna()**:

```{r prep_data, warning=FALSE, results="hide", message=FALSE,}
attach(transducin)
```

```{r nma.pdbs, cache=TRUE, warning=FALSE, results="hide", message=FALSE}
modes <- nma(pdbs, ncore=8)
```

```{r dccms, cache=TRUE}
cijs0 <- dccm(modes)
```

Below we perform optional filtering and averaging per state of the individual structure cross-correlation matrices.
In this example, we also utilize a cutoff for correlation (0.35 in this example) and a cutoff for C-alpha distance (10 angstrom) (See `help(dccm.filter)` and main text for details).
```{r dccm.filter, cache=TRUE, message=FALSE}
cij <- dccm.filter(cijs0, xyz=pdbs, fac=annotation[, "state3"], cutoff.cij=0.35, 
                dcut=10, scut=0, pcut=0.75, ncore=8)
```

```{r nets, cache=TRUE}
nets <- cna(cij, cutoff.cij = 0)
net1 = mod.select(nets$GTP)
net2 = mod.select(nets$GDI)
```

```{r layout2}
ref.pdb <- pdbs2pdb(pdbs, inds=1, rm.gaps=TRUE)[[1]]
cent.gtp.full = layout.cna(net1, pdb=ref.pdb, full=TRUE, k=3)[,1:2]
cent.gtp = layout.cna(net1, pdb=ref.pdb, k=3)[,1:2]
cent.gdi.full = layout.cna(net2, pdb=ref.pdb, full=TRUE, k=3)[,1:2]
cent.gdi = layout.cna(net2, pdb=ref.pdb, k=3)[,1:2]
```

```{r figure2, fig.cap="Correlation networks for GTP \"active\" and GDI \"inhibitory\" conformational states of transducin. Networks are derived from ensemble NMA of available GTP and GDI crystallographic structures.", echo=c(3,5,6,8)}
layout(matrix(c(1:4), 2, 2))
par(mar=c(0.1, 0.1, 3, 0.1))
plot.cna(net1, layout=cent.gtp.full, full=TRUE, vertex.label=NA, vertex.size=5)
title(main="GTP")
plot.cna(net1, layout=cent.gtp)
plot.cna(net2, layout=cent.gdi.full, full=TRUE, vertex.label=NA, vertex.size=5)
title(main="GDI")
plot.cna(net2, layout=cent.gdi)
```


# Document Details {.unnumbered}
This document is shipped with the Bio3D package in both R and PDF formats. All code can be extracted and automatically executed to generate Figures and/or the PDF with the following commands:

```{r eval=FALSE}
library(rmarkdown)
render("Bio3D_cna-transducin.Rmd", "all")
```


# Information About the Current Bio3D Session {.unnumbered}
```{r}
sessionInfo() 
```


# References {.unnumbered}


[^1]: The latest version of the package, full documentation and further
    vignettes (including detailed installation instructions) can be
    obtained from the main Bio3D website:
    [http://thegrantlab.org/bio3d/](http://thegrantlab.org/bio3d/).

[^2]: This vignette contains executable examples, see `help(vignette)`
    for further details.
